Sequential rank agreement methods for comparison of ranked lists

### 2015-06-17 latex/paper.Rnw
Tekst

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to improve the classification of acute leukemias
between two types: acute lymphoblastic leukemia (ALL) or acute myeloid
leukemia (AML) \citep{Golub1999}. Preprocessing of the gene expression
data was done as described in \citet{Dudoit2002} and the four
different analysis approaces were: marginal two-sample $t$ tests,
marginal logistic regression analyses, logistic regression eleastic
net, and marginal maximum information content correlations (MIC)
\citep{Reshef2011} which resulted in the rankings seen in
Table~\ref{tab1}.  The sequential rank agreement at depth $d$ can be
interpreted as how many ranks apart do we on average expect the items
found until depth $d$ to be so smaller values on the $y$ axis
correspond to \emph{better} agreement among the lists than larger
values.

@@

\end{center}
 \caption{Left panels: Sequential rank agreement for 4 different analysis methods
   applied to the 3051 genes in the Golub data. The two plots on the
   right show the sequential rank agreement for the same data but
   where only the top-\Sexpr{topk} ranked items are available. The blue
   and red areas correspond to the independent and randomized
   reference hypothesis areas, respectively. The bottom plots are
   identical to the corresponding top plots but have been zoomed in on
   the first part of the $x$ axis.}
 \label{fig:example1}
\end{figure}

If the lists generally agree on all items then we expect the
sequential rank agreement curve to be flat, but if there are changes
in the levels of rank agreement then this suggests that there are sets
of items that the lists agree on and other sets where they rank the
items differently. In particular, we generally expect the agreement
among the lists to be better towards the top of the lists and worse
towards the end of the lists, and in this case the sequential rank
agreement curve will start at a low level, and then have an increase
before it levels off.

@@

Discuss assumption

Present stuff that is long enough

information

in
\eqref{formula:censored-agreement}

Unless we have additional prior knowledge about the distribution of the
unranked items then we can restrict attention of the sequential rank
agreement to
\begin{equation}
d \leq \max(k_1, \ldots, k_L),
\end{equation}
since we have no information about the ordering of the items after a
depth $d$ where all lists are censored.

 ess contains full set of ranks and by using
the ... instead of applying the average ...

the mean then we

FIXME: If instead of running through all elements of $\tilde R_1(X_p)$
one would use the average rank in $\tilde R_1(X_p)$ we would end up with
a too small variance.

FIXME: For censored lists we expect a change point around the depth
where the lists are censored. This is an artefact stemming from the
fact that

@@

@@

<<>>=
library(changepoint)
cp1 <- cpt.mean(res, method="BinSeg")
cp2 <- cpt.mean(res2, method="BinSeg")
cp1@cpts
cp1@param.est
cp2@cpts
cp2@param.est
@

@@

\section{Comparison to ...}

In a classical paper by \citet{Golub1999} a dataset of 3051 gene
expression values were measured on 38 tumor mRNA samples in order to
improve the classification of acute leukemias between two types: acute
lymphoblastic leukemia (ALL) or acute myeloid leukemia
(AML). Preprocessing of the gene expression data was done as described
in \citep{Dudoit2002}.

We analyzed these gene expression data using four different
approaches: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net, and marginal
maximum information content correlations (MIC) \citep{Reshef2011}. For
the first two methods, the genes were ranked according to $p$ value.

<<fig-ao1, echo=FALSE, fig=TRUE,include=FALSE>>=
myres <- average_overlap(inputmatrix[,1:2])
myres2 <- average_overlap(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(myres, xlim=c(1, 300), ylim=c(0,1), type="l")
lines(xvals[-c(1,2)], myres2[1:498], col="red")
@

<<fig-ao2, echo=FALSE, fig=TRUE,include=FALSE>>=
mmyres <- sqrt(sra(inputmatrix[,1:2]))
mmyres2 <- sqrt(sra(inputmatrix[-c(1,2),1:2]))
xvals <- 1:500
plot(mmyres, xlim=c(1, 300), type="l")
lines(xvals[-c(1,2)], mmyres2[1:498], col="red")
@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig-ao1}
\includegraphics[width=.45\textwidth]{paper-fig-ao2}
\end{center}
 \caption{Sequential rank agreement for 4 different analysis methods
   applied to the Golub data (thick red line). The red and blue shaded
 areas represent the pointwise 95\% normal areas for the XXX and XXY benchmarks,
respectively.}
 \label{fig:case1}
\end{figure}

>>>>

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to improve the classification of acute leukemias
between two types: acute lymphoblastic leukemia (ALL) or acute myeloid
leukemia (AML) \citep{Golub1999}. Preprocessing of the gene expression
data was done as described in \citet{Dudoit2002} and the four
different analysis approaces were: marginal two-sample $t$ tests,
marginal logistic regression analyses, logistic regression eleastic
net, and marginal maximum information content correlations (MIC)
\citep{Reshef2011}. For the first two methods, the genes were ranked
according to $p$ value, for logistic regression the genes were ordered
by size of the corresponding coefficients (after standardization), and
MIC was ordered by correlation which resulted in the rankings seen in
Table~\ref{tab1}.  The sequential rank agreement at depth $d$ can be
interpreted as how many ranks apart do we on average expect the items
found until depth $d$ to be so smaller values on the $y$ axis
correspond to \emph{better} agreement among the lists than larger
values.

@@

\end{center}
 \caption{Left panels: Sequential rank agreement for 4 different analysis methods
   applied to the 3051 genes in the Golub data. Right panels:
   Corresponding sequential rank agreement for the same data but
   where only the top \Sexpr{topk} ranked items are available. The blue
   and red areas correspond to the independent and randomized
   reference hypothesis areas, respectively. The bottom plots are
   identical to the corresponding top plots but have been zoomed in on
   the first part of the $x$ axis.}
 \label{fig:example1}
\end{figure}

If the lists generally agree on all items then we expect the
sequential rank agreement curve to be flat, but if there are changes
in the levels of rank agreement then this suggests that there are sets
of items that the lists agree on and other sets where they rank the
items differently. In particular, we generally expect the agreement
among the lists to be better towards the top of the lists and worse
towards the end of the lists, and in this case the sequential rank
agreement curve will start at a low level and then increase until it
levels off.

@@

The proposed approach is based on two assumptions: 1) that the items
of interest will be found towards the top of the lists, and 2) that
the censored rankings provide so little information that it is
reasonable to assume that the missing rankings provide no information
and might as well be considered random.  The first assumption is
justifiable because we have already accepted that it is reasonable to
rank the items in the first place. The second assumption is fair
in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are combined then it is clear that the
interesting part of the sequential rank agreement curves must be for
smaller depths. Unless we have additional prior knowledge about the
distribution of the unranked items then we can restrict attention of
the sequential rank agreement to
\begin{equation}
d' \leq \max(k_1, \ldots, k_L),
\end{equation}
since we have no real information about the ordering of the items after a
depth $d'$ where all lists are censored.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For censored lists we also expect a
change point around the depth where the lists are censored. This is an
artefact stemming from the fact that we assume that the remainder of
the lists can be replaced by a simple permutation of the missing
items.

@@

\section{Comparison to average overlap}

<<fig-ao1, echo=FALSE, fig=TRUE,include=FALSE>>=
myres <- average_overlap(inputmatrix[,1:2])
myres2 <- average_overlap(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(xvals, myres[xvals], xlim=c(1, 150), ylim=c(0,1), type="l", lwd=2, ylab="Average overlap", xlab="List depth")
lines(xvals[-c(1,2)], myres2[1:498], col="red", lwd=2)
@

<<fig-ao2, echo=FALSE, fig=TRUE,include=FALSE>>=
mmyres <- sra(inputmatrix[,1:2])
mmyres2 <- sra(inputmatrix[-c(1,2),1:2])
xvals <- 1:500
plot(xvals, mmyres[xvals], xlim=c(1, 150), ylim=c(0, 110), type="l", lwd=2, ylab="Sequential rank agreement", xlab="List depth")
lines(xvals[-c(1,2)], mmyres2[1:498], col="red", lwd=2)
@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig-ao1}
\includegraphics[width=.45\textwidth]{paper-fig-ao2}
\end{center}
 \caption{Average overlap (left figure) and sequential rank agreement
   (right figure) for two different analysis methods
   applied to the Golub data (thick red line). The red and blue shaded
 areas represent the pointwise 95\% normal areas for the XXX and XXY benchmarks,
respectively.}
 \label{fig:case1}
\end{figure}

@@

<<>>=
#library(changepoint)
#cp1 <- cpt.mean(res, method="BinSeg")
#cp2 <- cpt.mean(res2, method="BinSeg")
#cp1@cpts
#cp1@param.est
#cp2@cpts
#cp2@param.est
@

@@
